\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-10 tex/paper.tex
minor notation changes

\usepackage{bm}
\usepackage{cancel}
\usepackage{bbold}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}

@@

@@

We consider a family of parameterized densities $p(\mathbf{x}|\theta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\theta$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p(\mathbf{x}|\theta) = \int p(\mathbf{x}|\mathbf{z},\theta) p(\mathbf{z}|\theta) d\mathbf{z}
\end{equation}
where $\mathbf{z} \in {\cal Z}$ is a latent variable providing an external
source of randomness. In particular, $\mathbf{z}$ is not necessarily assumed to
be a fixed-size vector (e.g., it can be a sequence of variable length) and its
distribution $p(\mathbf{z}|\theta)$ may itself depend on $\theta$ in some intricate way.

We assume that we already have an accurate simulation of the stochastic
generative process that defines $p(\mathbf{x}|\mathbf{z},\theta)$, as
specified through a highly regularized deterministic function $g(\cdot; \theta) : {\cal Z} \to
\mathbb{R}^d$ with usually few parameters. That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\theta) \equiv \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta)
\end{equation}
such that the likelihood $p(\mathbf{x}|\theta)$ can be rewritten as
\begin{equation}\label{eqn:p_x_sim}
    p(\mathbf{x}|\theta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p(\mathbf{z}|\theta) \mu(d\mathbf{z}),
\end{equation}
where $\mu$ is a probability measure.
Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
used to generate data in forward mode. For this reason, evaluating the integral
in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$,
e.g. as when $g$ is specified as a computer program.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is the inference of the parameters
of interest $\theta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the modeled data distribution $p(\mathbf{x}|\theta)$ induced by $g(\cdot;
\theta)$ over $\mathbf{z}$. That is,
\begin{equation}
    \theta^* = \arg \min_{\theta \in \Theta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\theta)),
\end{equation}
where $\rho$ is some distance or divergence.

@@

Generative adversarial networks (GANs) were first proposed by
\cite{goodfellow2014generative} as a way to build an implicit generative model
capable of producing samples from random noise $\mathbf{z}$. More specifically,
a generative model $g(\cdot; \theta)$ is pit against an adversarial
classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
from generated data $\tilde{\mathbf{x}} = g(\mathbf{z}; \theta)$. Both models $g$ and $d$
are trained simultaneously, in such a way that $g$ learns to fool
its adversary $d$ (which happens when $g$ produces samples comparable to the
observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
trained to optimality before each parameter update of the generator, it can
be shown that the original adversarial learning procedure~\cite{goodfellow2014generative} amounts to minimizing
the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\theta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\theta)$.

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W(p_r(\mathbf{x}), p(\mathbf{x}|\theta))$ by
replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. 
\begin{align}
   W( p_r(\mathbf{x} ), p(\mathbf{x} | \theta)  ) = \mathcal{L}_W = \sup_{f} \mathbb{E}_{p(\tilde{\mathbf{x}}|\theta)} [f(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [f(\mathbf{x})]
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\theta$ in order to respectively minimize

\begin{align}
    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - {\cal L}_W
\end{align}
%\begin{align}
%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]  \nonumber \\
%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)]
%\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] = U(\psi),
\end{equation}
where $q(\theta|\psi)$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\theta^* = \min_{\theta \in \Theta} f(\theta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\psi)$ is differentiable with respect to $\psi$, and using the log-likelihood
trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] \nonumber \\
    %&= \nabla_\psi \int f(\theta)  q(\theta|\psi)  d\theta \nonumber \\
    &= \int f(\theta) \nabla_\psi q(\theta|\psi)  d\theta \nonumber \\
    &= \int \left[ f(\theta) \nabla_\psi \log q(\theta|\psi) \right]  q(\theta|\psi) d\theta \nonumber \\
    &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta) \nabla_\psi \log q(\theta|\psi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log
q(\theta|\psi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\theta})$ for any $\theta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\psi)$ with stochastic gradient descent (or a variant thereof, like
Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W}, for scaling invariance and
robustness to noisy gradients).

% In this simple form, stochastic gradient descent based on
% Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
% of scale invariance. To avoid the dependence on the parameterization of the
% proposal distribution, natural gradients~\citep{amari1998natural} can be used
% instead to point towards the steepest descent in the space of realizable
% proposal distributions, rather than in the space of their parameters. Effectively,
% the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
% corrects the gradient of $U$ according to the local curvature of the KL-divergence
% surface of proposal distributions, which results in forming
% \begin{equation}\label{eqn:approx-grad-nat}
%     \tilde{\nabla}_\psi U(\psi) = \mathbf{F}(\psi)^{-1} \nabla_\psi U(\psi),
% \end{equation}
% where $\mathbf{F}(\psi)$ is the Fisher information matrix
% \begin{equation}
%     \mathbf{F}(\psi) = \mathbb{E}_{\theta \sim q_\phi}[ \nabla_\psi \log q_\psi(\theta) \nabla_\psi \log q_\psi(\theta)^{\top} ]
% \end{equation}
% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\psi \log q_\psi(\theta)$ already
% evaluated to form $\nabla_\psi U(\psi)$.

@@

    \begin{flushleft}
        {\it Inputs:} observed data $\{ \mathbf{x}_i \sim p_r(\mathbf{x}) \}_{i=1}^N$, simulator $g$.\\
        {\it Outputs:} proposal distribution $q(\theta|\psi)$, such that $p_r(\mathbf{x}) \approx p(\mathbf{x}|\psi)$.\\
        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the size $M$ of a mini-batch; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
    \end{flushleft}

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q(\theta|\psi) \leftarrow \text{prior on $\theta$ (with differentiable and known density)}$}
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r(\mathbf{x}), \theta_m \sim q(\theta|\psi), \mathbf{z}_m \sim p(\mathbf{z}|\theta_m), \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{Adam}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{ \theta_m \sim q(\theta|\psi),  \mathbf{z}_m \sim p(\mathbf{z}|\theta_m) \}_{m=1}^M$.}  \Comment{Update $q(\theta|\psi)$}
            \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
            % \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$}
            \State{$\psi \leftarrow \text{Adam}(\nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
GANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
the setting where we are interested in the inference of the parameters of a
fixed non-differentiable simulator (Section~\ref{sec:problem}),
rather than in learning the generative model itself,
gradients $\nabla_\theta g$ either do not exist or cannot be accessed. As a
result, gradients $\nabla_\theta {\cal L}_g$ cannot be constructed and the
optimization procedure cannot be carried out.

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\theta|\psi)$ over the
parameters of $g$ and $p(\mathbf{x}|\theta)$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$\phi$, unbiased estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$\psi$, estimates of $\nabla_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}\label{eqn:grad-ug-approx}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q(\theta_m|\psi)
\end{equation}
for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p(\mathbf{x}|\psi) \equiv \theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q(\theta|\psi)$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
More formally, the learned model  therefore corresponds to the marginal distribution
of the generated data. That is,
\begin{equation}
    p(\mathbf{x}|\psi) = \int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta.
\end{equation}

>>>>

\usepackage{bm}
\usepackage{cancel}
\usepackage{bbold}
\usepackage{bm}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}

@@

\newcommand{\qxpsi}{q(\mathbf{x}|\bfpsi)}
\newcommand{\bftheta}{{\bm \theta}}
\newcommand{\bfpsi}{{\bm \psi}}
\newcommand{\bfx}{\mathbf{x}}
\newcommand{\bfz}{\mathbf{z}}

@@

We consider a family of parameterized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
\end{equation}
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness. In particular, $\bfz$ is not necessarily assumed to
be a fixed-size vector (e.g., it can be a sequence of variable length) and its
distribution $p(\bfz|\bftheta)$ may itself depend on $\bftheta$ in some intricate way.

We assume that we already have an accurate simulation of the stochastic
generative process that defines $p(\mathbf{x}|\bfz,\bftheta)$, as
specified through a highly regularized deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
\mathbb{R}^d$ with usually few parameters. That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
\end{equation}
such that the likelihood $p(\mathbf{x}|\bftheta)$ can be 
considered as a non-differentiable reparameterization of the random variable $\bfz$.
%rewritten as
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
%where $\mu$ is a probability measure.
Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
used to generate data in forward mode. For this reason, evaluating the integral
in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
e.g. as when $g$ is specified as a computer program.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is the inference of the parameters
of interest $\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the modeled data distribution $p(\mathbf{x}|\bftheta)$ induced by $g(\cdot;
\bftheta)$ over $\bfz$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta \in \bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
\end{equation}
where $\rho$ is some distance or divergence.

@@

Generative adversarial networks (GANs) were first proposed by
\cite{goodfellow2014generative} as a way to build an implicit generative model
capable of producing samples from random noise $\bfz$. More specifically,
a generative model $g(\cdot; \bftheta)$ is pit against an adversarial
classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
from generated data $\tilde{\mathbf{x}} = g(\bfz; \bftheta)$. Both models $g$ and $d$
are trained simultaneously, in such a way that $g$ learns to fool
its adversary $d$ (which happens when $g$ produces samples comparable to the
observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
trained to optimality before each parameter update of the generator, it can
be shown that the original adversarial learning procedure~\cite{goodfellow2014generative} amounts to minimizing
the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\bftheta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\bftheta)$.

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta))$ by
replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. 
\begin{align}
   W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) = \mathcal{L}_W = \sup_{f} \mathbb{E}_{p(\tilde{\mathbf{x}}|\bftheta)} [f(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [f(\mathbf{x})]
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\bftheta$ in order to respectively minimize

\begin{align}
    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - {\cal L}_W
\end{align}
%\begin{align}
%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]  \nonumber \\
%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\phi)]
%\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\bftheta \in \bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \bftheta} f(\bftheta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
    &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
    &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, like
Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W}, for scaling invariance and
robustness to noisy gradients).

% In this simple form, stochastic gradient descent based on
% Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
% of scale invariance. To avoid the dependence on the parameterization of the
% proposal distribution, natural gradients~\citep{amari1998natural} can be used
% instead to point towards the steepest descent in the space of realizable
% proposal distributions, rather than in the space of their parameters. Effectively,
% the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
% corrects the gradient of $U$ according to the local curvature of the KL-divergence
% surface of proposal distributions, which results in forming
% \begin{equation}\label{eqn:approx-grad-nat}
%     \tilde{\nabla}_\bfpsi U(\bfpsi) = \mathbf{F}(\bfpsi)^{-1} \nabla_\bfpsi U(\bfpsi),
% \end{equation}
% where $\mathbf{F}(\bfpsi)$ is the Fisher information matrix
% \begin{equation}
%     \mathbf{F}(\bfpsi) = \mathbb{E}_{\bftheta \sim q_\phi}[ \nabla_\bfpsi \log q_\bfpsi(\bftheta) \nabla_\bfpsi \log q_\bfpsi(\bftheta)^{\top} ]
% \end{equation}
% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\bfpsi \log q_\bfpsi(\bftheta)$ already
% evaluated to form $\nabla_\bfpsi U(\bfpsi)$.

@@

    \begin{flushleft}
        {\it Inputs:} observed data $\{ \mathbf{x}_i \sim p_r(\mathbf{x}) \}_{i=1}^N$, simulator $g$.\\
        {\it Outputs:} proposal distribution $q(\bftheta|\bfpsi)$, such that $\qxpsi \approx p_r(\mathbf{x})$.\\
        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the size $M$ of a mini-batch; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
    \end{flushleft}

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q(\bftheta|\bfpsi) \leftarrow \text{prior on $\bftheta$ (with differentiable and known density)}$}
        \While{$\bfpsi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r(\mathbf{x}), \bftheta_m \sim q(\bftheta|\bfpsi), \bfz_m \sim p(\bfz|\bftheta_m), \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\bfz_m; \bftheta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{Adam}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{ \bftheta_m \sim q(\bftheta|\bfpsi),  \bfz_m \sim p(\bfz|\bftheta_m) \}_{m=1}^M$.}  \Comment{Update $q(\bftheta|\bfpsi)$}
            \State{$\nabla_\bfpsi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m)) \nabla_\bfpsi \log q_\bfpsi(\bftheta_m)$}
            % \State{$\mathbf{F}(\bfpsi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\bfpsi \log q_\bfpsi(\bftheta_m) \nabla_\bfpsi \log q_\bfpsi(\bftheta_m)^{\top}$}
            \State{$\nabla_\bfpsi H(q_\bfpsi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\bfpsi q_\bfpsi(\bftheta_m) \log q_\bfpsi(\bftheta_m)$}
            \State{$\bfpsi \leftarrow \text{Adam}(\nabla_\bfpsi U_g + \gamma \nabla_\bfpsi H(q_\bfpsi))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
GANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
the setting where we are interested in the inference of the parameters of a
fixed non-differentiable simulator (Section~\ref{sec:problem}),
rather than in learning the generative model itself,
gradients $\nabla_\bftheta g$ either do not exist or cannot be accessed. As a
result, gradients $\nabla_\bftheta {\cal L}_g$ cannot be constructed and the
optimization procedure cannot be carried out.

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of $g$ and $p(\mathbf{x}|\bftheta)$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\bfpsi$.
When updating
$\phi$, unbiased estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$\bfpsi$, estimates of $\nabla_\bfpsi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}\label{eqn:grad-ug-approx}
    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\phi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \phi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\bfpsi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
More formally, the learned model  therefore corresponds to the marginal distribution
of the generated data. That is,
\begin{equation}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
